1. Introduction
===============
The purpose of this project is to build a model that predicts the price of a diamond based on a number of variables. According to the project description:

_"This individual assignment asks students to use a training-and-test regimen to compare alternative regression methods, such as, traditional multiple regression, tree-structured regression, and neural networks."_

As such, the focus of this project will be on prediction accuracy and less on inferential studies.

2. Dataset
==========
This dataset consists of 7 variables. The target variable we are aiming for is __price__, a continuous integer. Our __6 explanatory variables__ are split into 2 categories: diamond characteristic and diamond vendor. The diamond characteristics are stated in the project description as the 4Cs: carat, color, clarity, and cut. In total there are __425 obervations__ in this dataset.

3. Analysis & Results
=====================
3.1 Data Preparation
---------------
### 3.1.1 Import Data
First we import the data via the provided CSV file.
```{r}
diamonds <- read.csv("two_months_salary.csv")
```
Let's take a quick look at the data:
```{r}
head(diamonds)
str(diamonds)
summary(diamonds)
```
We can see above that all __7 variables__ are present, the __datatypes have been set__ properly, and __there are no missing values__.

### 3.1.2 Data Modification
According to the provided analysis, it will be helpful to derive some variables initially. These variables specifically are __logprice__ (log transformation of the price variable) and __internet__ (whether or not the vendor is an online vendor). 
```{r}
diamonds$logprice <- log(diamonds$price)
diamonds$internet <- ifelse((diamonds$channel == "Internet"),2,1)
diamonds$internet <- factor(diamonds$internet,levels=c(1,2),labels=c("NO","YES"))
```
Let's take a look at the dataset again to verify the additions:
```{r}
str(diamonds)
```

3.2 Graphical Summary
---------------------
### 3.2.1 Provided Analysis
The graphical summary code was provided to us in advance. The code and the analysis are shown below:
```{r, fig.width=6, fig.height=6}
library(lattice)
xyplot(jitter(price) ~ jitter(carat) | channel + cut, 
       data = diamonds,
        aspect = 1, 
        layout = c(3, 2),
        strip=function(...) strip.default(..., style=1),
        xlab = "Size or Weight of Diamond (carats)", 
        ylab = "Price")
```
To summarize the analyses provided with the project:
- price and carat are numeric variables with a strong relationship
- cut and channel are factor variables related to price

### 3.2.2 Multivariate EDA on Price
It may also help to analyze each explanatory variable and its relationship to the target variable, price. For this portion we will be using the ggplot2 package.
```{r}
#library(ggplot2)
attach(diamonds) # attach dataset for easier selection
```

#### 3.2.2.1 Carat
```{r, fig.width=4, fig.height=4}
plot(carat, price)
```
It looks as if there's a __linear relationship between carat and price__. However, the price spreads out quite a bit as the carat number increases so this relationships is not for certain.

```{r}
detach(diamonds)
```

3.3 Modeling: Iteration 1
-------------------------
The first model we will apply is a mulitple regression. This method is quick, easy, and will provide us 